%!TEX root = ComputerScienceOne.tex

%Encapsulation - exercises

\section{Exercises}

\begin{exer}
A \emph{complex} number consists of two real numbers: a real
component $a$ and an imaginary component $bi$ where $b$ is a real number and
$i = \sqrt{-1}$.  Define an object or structure to model a
complex number.  Write functions to:
\begin{itemize}
  \item Create a complex number
  \item Print a complex number
  \item Perform basic arithmetic operations on two complex
  	numbers including addition, subtraction and multiplication 
	as defined by:
	$$c_1 + c_2 = (a_1 + b_1i) + (a_2 + b_1i) = (a_1+a_2) + (b_1+b_2)i$$
	$$c_1 - c_2 = (a_1 + b_1i) - (a_2 + b_1i) = (a_1-a_2) + (b_1-b_2)i$$
	$$c_1 \cdot c_2 = (a_1 + b_1i) \cdot (a_2 + b_1i) = 
	  (a_1a_1-b_1b_2) + (a_1b_2+b_1a_1)i$$
\end{itemize}
\end{exer}

\begin{exer}
\label{exercise:albumObject}
Design an object (or structure) that models an album.  Include
at least the album title, artist (or band) and release year.
Include any other data that you think is relevant and write
functions to support your object.
\end{exer}

\begin{exer}
\label{exercise:accountObject}
Design an object (or structure) that models a bank savings account.
Include at least the balance, APR, an account ``number'' and 
customer information (which may be another object or structure).
Include any other data that you think is relevant and write
functions to support your object.
\end{exer}

\begin{exer}
\label{exercise:stadiumObject}
Design an object (or structure) that models a sports stadium.
Include at least the stadium name, the team that plays there, 
its city, state, and year built.  Include both latitude and
longitude data.  Include any other data that you think is relevant 
and write functions to support your object.  Write a parser to
process a flat file data of all stadiums (in your chosen sport) 
and build instances of all of them.
\end{exer}

\begin{exer}
\label{exercise:networkObject}
Design an object (or structure) that models a network-connected
electronic device.  Include at least a unique ID, a human-readable
name for the device, a \gls{macLabel} address and \gls{ipLabel}
address as well as the device's bandwidth in megabits per second.
Include any other data that you think is relevant and write functions 
to support your object.
\end{exer}

\begin{exer}
\label{exercise:airportObject}
Design an object (or structure) that models an airport.
Include at least the name, FAA designation, 
its city, state, and latitude/longitude data.  Include any other 
data that you think is relevant and write functions to support your 
object.
\end{exer}


